---
title: "Assignment 9: Class Wrap-up"
format: html
editor: visual
author: "CS-201 Students, Fall 2025"
date: "`r Sys.Date()`"
toc: true
---

We are finally at the end of the semester! The final assignment for the class will be to summarize and share some of the things we've done so far.

The plan is for everyone to be randomly paired with a tiny topic pertaining to something we've learned together this semester and then submit a code snippet to be presented in class on Wednesday, 11/12. (Think \~2 minute walkthrough)

#### Libraries

```{r, message=FALSE}
library(tidyverse)
```

#### Read in the list of topics

```{r, message=FALSE}
# topics <- read_csv("data/cs201_topics.csv")
# view(topics)

# anything to edit?
# topics[7, "topic"] <- "wildcard"
```

#### Read in the list of student names

```{r, message=FALSE}
# names <- read_csv("data/cs201_names.csv")
# view(names)
```

#### Generate a random number in the names dataframe

```{r, message=FALSE}
# names <- names %>%
#   mutate(number = sample(1:nrow(topics), nrow(names), replace = FALSE))
# 
# view(names)
```

#### Join the names to the topics

```{r, message=FALSE}
# assigned_topics <- names %>%
#   full_join(topics, by = "number")
# 
# print(assigned_topics)
# 
# write.csv(assigned_topics, "data/assignment9_assigned_topics.csv")

# note that if you run the above script and randomize the names again, your outcome will be different from mine!!
# go with the topics assigned in class and announced on Canvas!

assigned <- read_csv("data/assignment9_assigned_topics.csv")
assigned
```

## Instructions

1.  Find your assigned topic. Talk to me if you reeeeeeaaaally want to do something else, or trade with someone (but let me know).

2.  Prepare a code snippet that briefly (\~2) explains the topic in the code chunk below, using `penguins` data.

3.  Push to GitHub **BY 10:00am** on Wednesday 11/12 (I'll need time to compile before class).

4.  Come to class Wednesday and show us how it's done :)

#### Extra libraries needed (if any)

```{r, message=FALSE}
library(janitor)
library(here)
library(plotly)
```

#### Read in penguins data

```{r, message=FALSE}
penguins <- read_csv(here("data/penguins_lter.csv"))

penguins <- penguins %>% 
  clean_names()
```

## Tallen - boxplots vs. violin plots

```{r, message=FALSE, warning=FALSE}
ggplot(data = penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot() +
  geom_jitter(size = 0.5) + # adding some datapoints and jittering them
  theme_bw() + # sets overall plot theme (think background, axes, bounding box)
  scale_fill_viridis_d() +
  labs(title = "Palmer Penguins, Distribution of Flipper Length",
       subtitle = "3 penguin species observed on three islands in the Palmer Archipelago, Antarctica",
       x = "",
       y = "Flipper length (mm)") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), legend.position = "none") +
  coord_flip()

```

```{r, message=FALSE, warning=FALSE}
# Basic violin plot: body mass by species
ggplot(data = penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_violin(trim = FALSE) +
  labs(
    title = "Distribution of Penguin Body Mass by Species",
    x = "Penguin Species",
    y = "Body Mass (g)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")

```

## Brandon - `dplyr::filter`

```{r, message=FALSE, warning=FALSE}
head(penguins)

# Example 1: Filter for a single condition
# Show only penguins from the Adelie species
adelie_penguins <- penguins %>%
  filter(species == "Adelie Penguin (Pygoscelis adeliae)")

# Example 2: Filter for multiple conditions
# Show only female Chinstrap penguins with body mass over 3500g
chinstrap_males_heavy <- penguins %>%
  filter(species == "Chinstrap penguin (Pygoscelis antarctica)", sex == "MALE", body_mass_g > 3500)

# Example 3: Use OR conditions with |
# Show penguins that are either Adelie OR have a flipper length over 190 mm
adelie_or_long_bill <- penguins %>%
  filter(species == "Adelie Penguin (Pygoscelis adeliae)" | flipper_length_mm > 190)

# Print results
adelie_or_long_bill

```

## Aichen - `dplyr::mutate`

```{r, message=FALSE}
# insert code snippet here

```

## Calvin - `dplyr::summarize`

```{r, message=FALSE}
# insert code snippet here

```

## Amara - `dplyr::case_when` vs. `ifelse()`

```{r, message=FALSE}
# insert code snippet here

```

## Cody - `select()` and `rename()`

```{r, message=FALSE}
# insert code snippet here

```

## Kanani - `arrange()`

```{r, message=FALSE}
# insert code snippet here

```

## Grace - `facet_wrap()` vs. `facet_grid()`

```{r, message=FALSE}
# insert code snippet here

```

## Sophia - scatterplots and aesthetics (`aes`, `color`, `shape`, `size`)

-   `aes`: maps data to visuals

-   `color`: controls point color

-   `shape`: controls point shape/symbol

-   `size`: controls point size

```{r, message=FALSE, warning=FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
   geom_point(shape = 17) +
  theme_bw() +
  scale_color_viridis_d()

```

## Karl - customizing plots (labs, themes, titles)

#### labs/titles: adds titles, subtitles, and axis labels.

```{r, message=FALSE, warning=FALSE}
ggplot(data = penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(alpha = 0.5) +
  theme_classic() +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  labs(title = "Palmer Penguins, Distribution of Flipper Length",
       subtitle = "3 penguin species observed on three islands in the Palmer Archipelago, Antarctica",
       x = "Flipper Length (mm)",
       y = "Number of Penguins")

```

#### themes: applies a black and white theme or background.

There are other themes like theme_minimal(), classic() and more.

```{r, message=FALSE, warning=FALSE}
ggplot(data = penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(alpha = 0.5, bins = 30) +
  labs(
    title = "Palmer Penguins: Distribution of Flipper Length",
    subtitle = "Three species observed in the Palmer Archipelago, Antarctica",
    x = "Flipper Length (mm)",
    y = "Number of Penguins"
  ) +
  theme_linedraw()
```

#### fill: scale_fill_manual(): manually pick colors for values.

In the penguins case its species and it matches the penguins species category. You can change the colors to whatever color you want.

```{r, message=FALSE, warning=FALSE}
ggplot(data = penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(alpha = 0.5, bins = 30) +
  theme_dark() +
  scale_fill_manual(values = c("pink", "yellow", "cyan")) +
  labs(
    title = "Palmer Penguins: Distribution of Flipper Length",
    subtitle = "Three species observed in the Palmer Archipelago, Antarctica",
    x = "Flipper Length (mm)",
    y = "Number of Penguins"
  )
```

## Shelley - `plotly`

```{r, message=FALSE, warning=FALSE}
p <- ggplot(penguins, aes(x = culmen_length_mm, 
                          y = culmen_depth_mm, 
                          color = species)) +
  geom_point(size = 3) +
  labs( title = "Penguin Bill Size Comparison",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)") +
  theme_minimal()

ggplotly(p)

```

## Sano - reading in data (various options)

```{r, message=FALSE}
# insert code snippet here

```

## Sadie - `skimr` and/or `naniar`

```{r, message=FALSE}
# insert code snippet here

```

## Vlad - `count()` and `group_by()`

```{r, message=FALSE}
# insert code snippet here

```

## Johnny - `ggsave()` and `write.csv()`

Start by making a scatter plot of penguin body mass vs flipper length:

```{r, message=FALSE, warning=FALSE}
penguin_plot <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.8) +
  theme_bw() +
  labs(
    title = "Penguin Body Mass vs Flipper Length",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  )

penguin_plot
```

#### Save The Plot as Image File

`ggsave` saves the plot as a png in data folder

```{r, warning=FALSE}
ggsave("data/penguin_plot.png", penguin_plot, width = 6, height = 4, dpi = 300)
```

#### Summary Table of Average Body Mass/ Species

Create a summary table - groups data by species and calculate average body mass for each group.

```{r}
penguin_summary <- penguins %>%
  group_by(species) %>%
  summarise(avg_mass_g = mean(body_mass_g, na.rm = TRUE))

penguin_summary
```

#### Save Summary Table as CSV file

`write.csv` saves that summary table as csv file in the data folder

```{r}
write.csv(penguin_summary, "data/penguin_summary.csv", row.names = FALSE)
```

## Rei - wild card!

```{r, message=FALSE}
# insert code snippet here

```

## Larchelle - wild card!

```{r, message=FALSE}
# insert code snippet here

```
